\textcolor{blue}{Change Title} The Impact of Model Assumptions on RES Drought Analysis

### 2024-11-21 Article/main.tex
Update Article
Minor changes made, larger changes flagged.

%opening
\title{The Impact of Model Assumptions on RES Drought Analysis}
\author{Boris Morin, Damian Flynn, Aina Maimó Far, Conor Sweeney}

@@

@@

\textcolor{magenta}{Some things to keep in mind when reading from Damian's comments:
	1. The whole current/projected and existing/balanced thing. Pick one and stick to it, including the figures (also fix the a/c level in fig 10)
	2. See if you are happy with the section titles throughout the document}

@@

As the global shift towards renewable energy accelerates, the European Union (EU) has positioned itself as a leader in climate and energy policy. Under the revised European Green Deal \cite{greendeal2023report}, the EU has set ambitious targets to achieve at least 69\% of electricity generation from renewable energy sources (RES) by 2030. This is a significant increase from 2022, when RES accounted for over 41\% of the EU’s electricity generation \cite{europe2023stat}. Meeting this goal is a key step in reducing greenhouse gas emissions and ensuring a sustainable energy future.

However, as reliance on RES grows, so does the challenge of managing the variability of weather-dependent energy sources like wind and PhotoVoltaic (PV) power. This challenge is further intensified by the increasing electrification of energy sectors which places additional demand on the power system~\cite{bloomfield2021}. As more sectors shift from fossil fuels to electricity, the power grid becomes increasingly sensitive to meteorological conditions~\cite{bloomfield2016, vanderWiel2019drought}. In this context, periods of low renewable generation often referred to as \textit{Dunkelflaute} or RES droughts, pose a greater risk to grid stability and energy security. Ensuring a reliable electricity supply during these RES droughts is crucial for maintaining a resilient energy system capable of meeting both the growing demand for electricity and the EU’s decarbonization targets.

For this study, a RES drought event is defined when the average capacity factor (CF) remains below a fixed threshold for a given duration, following the methodology established in previous research~\cite{kaspar2019drought, ohba2022drought, mockert2023drought, mayer2023drought}. Other studies have employed varying methodologies for defining RES droughts. One approach uses relative thresholds that change over the course of the year to account for seasonal variations in renewable energy generation~\cite{raynaud2018drought, rinaldi2021drought, gangopadhyay2022drought, allen2023drought, kapica2024drought}. Another common method relies on quantile-based thresholds, where drought events are defined by identifying periods of unusually low generation relative to historical production levels, typically based on the lowest production percentiles~\cite{bracken2024drought, allen2023drought}. Additionally, some studies combine these definitions with metrics that incorporate the demand side of energy consumption, analysing the balance between supply and demand during drought periods~\cite{raynaud2018drought, rinaldi2021drought, allen2023drought, bracken2024drought}. In this paper, the focus is exclusively on energy generation, using a fixed threshold approach to define RES droughts.

RES drought events can be identified from wind and PV CF time series. In this study, we use three different datasets, all of them driven by ERA5 data~\cite{hersbach2020era5}. The first two datasets are part of C3S Energy (C3S-E)~\cite{cds2023energy}, an energy-based operational dataset produced by the EU Copernicus Climate Change Service~\cite{dubus2023energy} (C3S). One of the C3S-E datasets provides CF time series aggregated at the national scale. Another C3S-E dataset provides the CF time series at each grid point, at the ERA5 resolution of 0.25°. The third dataset is produced by the authors using the Atlite model~\cite{hofman2021atlite}, which converts the  ERA5 atmospheric data to a generation time series using specified wind turbine and PV panel models.

This study aims to quantify the occurrence of RES drought events in Ireland for two energy scenarios. The first represents the existing energy mix, predominantly driven by wind, while the second is based on a projected energy scenarios, featuring a more balanced share between wind and PV. A uniform methodology is used to identify RES droughts for each of the three datasets, and differences in the frequency, the duration and the seasonality are presented.

@@

This study uses publicly available datasets to construct and validate the models for estimating the CF of wind and PV energy. The primary data sources include EirGrid, the transmission system operator for the ic of Ireland, and the ERA5 reanalysis dataset.

@@

EirGrid, which operates in the Republic of Ireland, and SONI, the system operator for Northern Ireland, provide detailed datasets on all wind and PV farms across the island of Ireland (Republic of Ireland and Northern Ireland) from 1990 to the present~\cite{eirgrid}. These datasets include critical information such as each farm’s installed capacity, name, and connection date. To enhance the accuracy of this data, the longitude and latitude for each farm were manually determined through online searches.

The spreadsheet available from the EirGrid website contains two key variables: generation and availability values for wind (from 2014 onward) and PV (from 2018 onward). For this study, availability data, which indicates the amount of energy being delivered to the grid, is used. This value is combined with installed capacity to compute the CF, the main metric for assessing renewable energy output.

Detailed data analysis, supported by access to additional data from EirGrid, revealed that the PV installed capacity available on the spreadsheets is the Maximum Export Capacity (MEC) and does not accurately reflect the true generating capacity. Instead, MEC values from the spreadsheet are used as a basis. To adjust for actual grid conditions, installed capacities are assumed to be 1.4 times the MEC values. Whenever farm generation is calculated, this scaled capacity is used and any farm output is capped to the MEC.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables used in the models to generate the Atlite and C3S-Energy datasets.

\begin{table}[h!]
	\centering
	\begin{tabular}{|l|c|}
		\hline
		{\textbf{ERA5 name}}      & \textbf{variable} \\ \hline
		100 metre zonal and meridional wind speed   & $u_{100}$, $v_{100}$ \\
		2 metre temperature                         & $t2m$ \\
		Surface net solar radiation                 & $ssr$ \\
		Surface solar radiation downwards           & $ssrd$  \\
		Top of atmosphere incident radiation        & $tisr$  \\
		Total sky direct solar radiation at surface & $fdir$  \\ \hline
	\end{tabular}
	\caption{Variables retrieved from ERA5 dataset}
	\label{tab:var_name}
\end{table}

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. This choice is based on expert advice and actual trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: global horizontal irradiance (referred to here as $ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles based on the geographical location~\cite{saintdrenan2018solar}.

@@

In this study, we use three datasets to analyse RES droughts. Data downloaded from C3S-E are used to obtain two datasets: one based on national-level (C3S-E N), and another on grid-level (C3S-E G). The third dataset is computed using the Atlite model (Atlite).

@@

For national-level analyses, we use the aggregated CF time series provided by C3S-E at two levels: Republic of Ireland (NUTS0: IE) and Northern Ireland (NUTS2: UKN0). We compute a weighted average of these, based on the installed capacity of each one, to represent the total CF of Ireland. This CF time series for Ireland is based on the underlying C3S-E assumption of RES generation at every 0.25\textdegree grid point in Ireland.

@@

Atlite is an open-source tool developed by PyPSA~\cite{hofman2021atlite}, and is widely used for estimating wind and PV generation~\cite{mockert2023drought, li2023atlite, parzen2023atlite, ali2023comparative}. This model transforms weather data into energy data, using the locations of existing RES farms as described in C3S-E G. A key distinction between C3S-E and Atlite lies in their representation of wind turbines and PV panels.

For wind power estimation, Atlite uses the wind speed at 100 metres ($u_{100}$, $v_{100}$). We provide Atlite with a wind turbine power curve generated by Renewables.ninja: Enercon E112.4500 with 0.30w smoothing~\cite{staffell2016wake} at a hub height of 100 metres, which it uses when converting wind speed to CF. For PV power estimation, Atlite uses four irradiance variables ($ssr$, $ssrd$, $tisr$, and $fdir$) as well as the air temperature ($t2m$). We use a single PV model~\cite{beyer2004pv}, with the Kaneka Hybrid panel option. The azimuth angle is fixed at 180\textdegree (due south) and we select the optimal tilt angle option.

@@

This study investigates wind and PV capacity factors (CF) using three different models, resulting in six individual CF time series based on the installed capacities of 2023. In addition to analysing wind and PV generation separately, a combined CF is computed for each model by averaging wind and PV generation, weighted by their installed capacities at the end of 2023 (6.7 GW for wind and 0.6 GW for PV). This results in a total of nine CF time series for the existing scenario, which represents the existing distribution of RES installations in Ireland.

Given that PV capacity in Ireland remains relatively low in 2023, and to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considers a balanced scenario. In this scenario, the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW, based on targets outlined in the roadmap published by EirGrid and SONI~\cite{eirgrid2023future}. For PV, new time series are generated for both the Atlite and C3S-E G models, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap. For wind, the CF time series remain unchanged, as significant shifts in wind farm distribution are not expected. The PV and wind time series are then combined using updated weights that reflect the installed capacities in the balanced scenario (9 GW for wind and 8 GW for PV).

@@

This approach allows for a comparative analysis between a wind-dominated system (the existing scenario) and a more evenly distributed system (the balanced scenario), assessing how the balance of RES capacity affects the occurrence of RES droughts. The results for these two scenarios are analysed and discussed in the following sections.

@@

In this study, a RES drought event is defined as occurring when the 24-hour moving average of CF remains below a fixed threshold of 0.1 for a period longer than 24 hours. The choice of this threshold is somewhat arbitrary but aligns with similar studies on low renewable energy production \cite{kaspar2019drought, ohba2022drought, mayer2023drought} . By using a 24-hour moving average, fewer but longer-lasting events are captured compared to using the raw CF time series, which can be more sensitive to short-term fluctuations. Although other methods exist for identifying RES droughts, a fixed threshold has been chosen in this study to enable consistent inter-comparison across the three datasets.

@@

Fig.~\ref{fig:find_res_droughts} illustrates the moving average identification method. When applying the moving average, this CF time series identifies a single event lasting 161 hours. In contrast, using the raw CF series would have resulted in three separate events, with durations of 36, 35, and 40 hours.

The moving average approach smooths out short-term fluctuations, so that brief periods above the threshold do not interrupt an otherwise continuous low-CF period. This means that a single hour above the threshold does not "break" a drought event if it is surrounded by prolonged low-generation hours. As a result, fewer but longer-lasting drought events are identified, which may better reflect real-world conditions where energy supply constraints persist over extended periods.

@@

Before going into the analysis of RES droughts, it is important to first verify the accuracy of the datasets used in this study. For the verification process, we use time-varying values of installed capacity to account for changes in RES development over the verification period. This step allows us to assess how well the datasets represent the production of renewable energy by comparing them against observed data.

@@

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve}a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provide four associated smoothed power curves. The smoothing is done by using a Gaussian filter with different standard deviations that depend on the wind speed. A wind CF time series for Ireland for all wind turbine power curve options and all smoothings is generated. The performance of each wind turbine power curve is then calculated based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$  smoothing filter.

@@

The smoothing of the wind turbine power curve represents losses associated with each individual turbine and the interaction between the different turbines, which is important when modelling wind energy on larger scales. The histogram in Fig.~\ref{fig:power_curve}b shows that the C3S-E power curve tends to underestimate low CF values and overestimate higher ones, whereas the smoothed Atlite power curve more closely follows the recorded wind availability data from EirGrid.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_wind_contour}
	\caption{Wind CF density plot of the observed EirGrid CF (vertical axes) and modelled (horizontal axes) CF data for the Atlite  (left panel), C3S-E G (central panel) and C3S-E N (right panel) models}
	\label{fig:wind_verification_contour}
\end{figure}

The effect of the differences in the power curves is also visible in Fig.~\ref{fig:wind_verification_contour} where the Atlite model shows good agreement with the EirGrid data. In contrast, the two C3S-E datasets tend to overestimate the EirGrid CF. This is confirmed by Table~\ref{tab:wind_skill_scores}, which shows that Atlite performs better than the C3S-E datasets for all three skill scores.

@@

The C3S-E datasets calculate PV generation multiple times using the same model with varying azimuth and tilt angles. The results are then aggregated based on a statistical distribution linked to each geographical location. Similarly to the wind CF verification, the Atlite model allows the user to select a given PV panel characteristics. In this study we have chosen to look at the three PV panels offered by in the Atlite model. Following the same methodology as in the previous section, we compare the three available models using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, we selected the Breyer PV panel model \cite{beyer2004pv}, using the Kaneka Hybrid panel option.

@@

Fig.~\ref{fig:bar_number_events_verification_pv} shows the number of PV drought events in 2023 across different duration ranges. The figure reveals partial agreement between the three datasets and the EirGrid data, with consistent results observed for duration ranges of 1-2, 3-4, 7-8, and 8+ days. However, discrepancies appear in the other ranges, where the models diverge from the EirGrid data. The main challenge in validating PV data stems from the recent installation of a large share of Ireland’s PV capacity, leading to uncertainties in PV generation data and the actual generating capacity in the first few months after each farm is connected. With over 60\% of the total PV capacity installed in 2023, these data uncertainties significantly impact the ability to perform rigorous validation for PV drought events. While the results are reasonably aligned in some ranges, the inherent challenges in PV data highlight the limitations of this verification.

@@

In this section, we evaluate RES drought events under two different scenarios with fixed installed capacities: the existing scenario, with 89\% of wind capacity (6.7 GW) and 11\% of PV capacity (0.6 GW), and a balanced scenario, where wind capacity comprises 53\% (9 GW) and PV capacity increases to 47\% (8 GW). Both scenarios are driven by 45 years of ERA5 data. Using the RES drought identification process described earlier (Section~\ref{sec:res_drought}), we first analyse wind and PV droughts separately before presenting the results for combined (wind + PV) RES droughts.

@@

The bottom panels show the combination of wind and PV under the  two capacity scenarios. In the existing RES installation scenario (Fig.~\ref{fig:boxplot_number_events}c), the identified RES droughts closely match those for wind alone, which is expected due to the dominance of installed wind capacity. In contrast, the balanced scenario (Fig.~\ref{fig:boxplot_number_events}d) shows a clear reduction in the number of drought events across all datasets and durations, with a decrease of the total number of events of 56\% for Atlite, 52\% for C3S-E G, and 50\% for C3S-E N. This reduction is attributed to the anti-correlation between wind and PV production throughout the year, a phenomenon well-documented in the literature \cite{kaspar2019drought} and discussed in further detail in the later analysis.

@@

The RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period refers to the estimated average interval between occurrences of events with a specified duration or intensity. Fig.~\ref{fig:return_periods} illustrates the return periods for varying RES drought durations, highlighting how often different drought lengths are likely to occur across the datasets. This analysis provides insight into the frequency and likelihood of prolonged low-generation periods, which is crucial for evaluating the potential impact of RES droughts on energy reliability.

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves significantly different than the two C3S-E datasets. The Atlite results follow the ones for wind with a log-linear increase, but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets likely arises from how each model handles PV output near the threshold. Atlite appears to hover slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.

For the existing installed capacity scenario (Fig.\ref{fig:return_periods}c), the return periods mirror those of wind (Fig.\ref{fig:return_periods}a), reinforcing that wind is the primary driver of drought patterns in the existing energy mix. In the balanced scenario (Fig.~\ref{fig:return_periods}d), the return periods for RES droughts increase across all durations. For example, the return period for a five-day drought event extends from roughly six months to four years in the Atlite dataset, and from about fifteen months to around five years in the two C3S-E datasets. The return periods for Atlite align more closely with those of C3S-E in the balanced scenario compared to the existing one. This convergence is due to the reduced proportion of wind capacity in the balanced scenario, which lessens the impact of wind-specific characteristics. Notably, there is a greater difference in how each model represents wind generation than PV, meaning the balanced scenario reduces these wind-related discrepancies, bringing the datasets closer together.

@@

Overall, the four panels indicate that regardless of the threshold applied, the Atlite dataset consistently identifies more events than the two C3S-E datasets, except for PV when the threshold is below 0.08. This finding suggests that Atlite, likely due to its more accurate representation of wind patterns in Ireland, captures more RES drought events that might be missed with different modeling assumptions in the C3S-E datasets.

>>>>

%opening
\title{ \textcolor{blue}{Change Title} The Impact of Model Assumptions on RES Drought Analysis}
\author{Boris Morin, Damian Flynn, Aina Maimó Far, Conor Sweeney}

@@

\textcolor{blue}{Add Abstract}

@@

@@

\textcolor{blue}{Get rid of ``we" everywhere.}

\textcolor{blue}{Make sure past tense is used everywhere, e.g. ``was", not ``is"}

\textcolor{blue}{Merge and shorten first two paragraphs} As the global shift towards renewable energy accelerates, the European Union (EU) has positioned itself as a leader in climate and energy policy. Under the revised European Green Deal \cite{greendeal2023report}, the EU has set ambitious targets to achieve at least 69\% of electricity generation from renewable energy sources (RES) by 2030. This is a significant increase from 2022, when RES accounted for over 41\% of the EU’s electricity generation \cite{europe2023stat}. Meeting this goal is a key step in reducing greenhouse gas emissions and ensuring a sustainable energy future.

However, as reliance on RES grows, so does the challenge of managing the variability of weather-dependent energy sources like wind and photovoltaic (PV) power. This challenge is further intensified by the increasing electrification of energy sectors which places additional demand on the power system~\cite{bloomfield2021}. As more sectors shift from fossil fuels to electricity, the power grid becomes increasingly sensitive to meteorological conditions~\cite{bloomfield2016, vanderWiel2019drought}. In this context, periods of low renewable generation, often referred to as \textit{Dunkelflaute} or RES droughts, pose a greater risk to grid stability and energy security. Ensuring a reliable electricity supply during these RES droughts is crucial for maintaining a resilient energy system capable of meeting both the growing demand for electricity and the EU’s decarbonization targets.

\textcolor{blue}{Insert sales pitch paragraph here}

For this study, a RES drought event is defined as occurring when the average capacity factor (CF) remains below a fixed threshold for a given duration, following the methodology used in other research~\cite{kaspar2019drought, ohba2022drought, mockert2023drought, mayer2023drought}. Alternative methodologies exist for defining RES droughts. One approach uses relative CF thresholds that change over the course of the year to account for seasonal variations in renewable energy generation~\cite{raynaud2018drought, rinaldi2021drought, gangopadhyay2022drought, allen2023drought, kapica2024drought}. Another common method relies on percentile-based thresholds, where drought events are defined by identifying periods of unusually low generation relative to historical production levels, typically based on the lowest production percentiles~\cite{bracken2024drought, allen2023drought}. Additionally, some studies combine these definitions with metrics that incorporate the demand side of energy consumption, analysing the balance between supply and demand during drought periods~\cite{raynaud2018drought, rinaldi2021drought, allen2023drought, bracken2024drought}. In this paper, the focus is exclusively on energy generation, and a fixed threshold approach to define RES droughts is used.

RES drought are identified using wind and PV CF time series. In this study, three different datasets are used, all of which are driven by ERA5 data~\cite{hersbach2020era5}. Two of the datasets are part of C3S Energy (C3S-E)~\cite{cds2023energy}, an energy-based operational dataset produced by the EU Copernicus Climate Change Service~\cite{dubus2023energy} (C3S). One of the C3S-E datasets used provides CF time series aggregated at the national scale, while the other provides the CF time series at each grid point, at the ERA5 resolution of 0.25°. The third dataset used was generated using the Atlite model~\cite{hofman2021atlite}, which converts the  ERA5 atmospheric data to a generation time series using specified wind turbine and PV panel models. Atlite is an open-source tool developed by PyPSA~\cite{hofman2021atlite}, and is widely used for estimating wind and PV generation~\cite{mockert2023drought, li2023atlite, parzen2023atlite, ali2023comparative}.

This study aims to quantify the occurrence of RES drought events in Ireland for two energy scenarios. The first represents the existing energy mix, predominantly driven by wind, while the second is based on a projected future energy scenario, featuring a more balanced share between wind and PV. A uniform methodology is used to identify RES droughts for each of the three datasets, and differences in the frequency, duration and seasonality are presented.

@@

This study uses publicly available datasets to construct and validate the models for estimating the CF of wind and PV energy. The primary data sources include: EirGrid, the transmission system operator for the island of Ireland; the ERA5 reanalysis dataset; and the C3S-E datasets.

@@

EirGrid, which operates in the Republic of Ireland, and SONI, the system operator for Northern Ireland, provide detailed datasets on all wind and PV farms across the island of Ireland (Republic of Ireland and Northern Ireland) from 1990 to the present~\cite{eirgrid}. These datasets include information such as each farm’s installed capacity, name, and connection date. To enhance the accuracy of this data, the longitude and latitude for each farm were manually determined through online searches.

\textcolor{blue}{Rewrite this paragraph:} The spreadsheet available from the EirGrid website contains two key variables: generation and availability values for wind (from 2014 onward) and PV (from 2018 onward). For this study, availability data, which indicates the amount of energy being delivered to the grid, is used. This value is combined with installed capacity to compute the CF, the main metric for assessing renewable energy output.

The PV installed capacity available on the spreadsheets from EirGrid represents the Maximum Export Capacity (MEC) and does not accurately reflect the installed PV capacity. To adjust for actual PV generation potential, installed capacities are estimated to be 1.4 times the MEC values. This installed capacity is used for PV genreation, with farm output capped to the MEC value.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables which have been used by Atlite and C3S-Energy.

\begin{table}[h!]
	\centering
	\begin{tabular}{|l|c|}
		\hline
		{\textbf{ERA5 name}}      & \textbf{variable} \\ \hline
		100 metre zonal and meridional wind speed   & $u_{100}$, $v_{100}$ \\
		2 metre temperature                         & $t2m$ \\
		Surface net solar radiation                 & $ssr$ \\
		Surface solar radiation downwards           & $ssrd$  \\
		Top of atmosphere incident radiation        & $tisr$  \\
		Total sky direct solar radiation at surface & $fdir$  \\ \hline
	\end{tabular}
	\caption{ERA5 variables used to calculate wind and PV generation.}
	\label{tab:var_name}
\end{table}

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. This choice is based on expert advice and the trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: Surface solar radiation downwards ($ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times, using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles based on the geographical location~\cite{saintdrenan2018solar}.

@@

In this study, we use three datasets to analyse RES droughts for teh island of Ireland. Data downloaded from C3S-E are used to obtain two datasets: one based on national-level data (C3S-E N), and another on grid-level data (C3S-E G). The third dataset is computed using the Atlite model (Atlite).

@@

\textcolor{blue}{Rewrite:} For national-level analyses, we use the aggregated CF time series provided by C3S-E at two levels: Republic of Ireland (NUTS0: IE) and Northern Ireland (NUTS2: UKN0). We compute a weighted average of these, based on the installed capacity of each one, to represent the total CF of Ireland. This CF time series for Ireland is based on the underlying C3S-E assumption of RES generation at every 0.25\textdegree grid point in Ireland.

@@

Atlite transforms weather data into energy data, using the wind speed at 100 metres ($u_{100}$, $v_{100}$), and the locations of existing RES farms as described in C3S-E G. A key distinction between C3S-E and Atlite lies in their representation of wind turbines and PV panels.

This study uses a wind turbine power curve generated by Renewables.ninja: Enercon E112.4500 with 0.30w smoothing~\cite{staffell2016wake}. For PV power estimation, Atlite uses four irradiance variables ($ssr$, $ssrd$, $tisr$, and $fdir$) as well as the air temperature ($t2m$). We use a single PV model~\cite{beyer2004pv}, with the Kaneka Hybrid panel option. The azimuth angle is fixed at 180\textdegree (due south) and we select the optimal tilt angle option.

@@

In addition to analysing wind and PV generation separately, a combined CF is computed for each model by averaging wind and PV generation, weighted by their installed capacities at the end of 2023 (6.7 GW for wind and 0.6 GW for PV). Given that PV capacity in Ireland remains relatively low in 2023, and to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considers a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These values are based on targets outlined in the roadmap published by EirGrid and SONI~\cite{eirgrid2023future}. New time series are generated for both the Atlite and C3S-E G PV models, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap. For wind, the CF time series remain unchanged, as significant shifts in wind farm distribution are not expected. The PV and wind time series are then combined using updated weights that reflect the installed capacities in the balanced scenario (9 GW for wind and 8 GW for PV).

@@

This approach allows for a comparative analysis between a wind-dominated system (the existing scenario) and a more evenly distributed system (the balanced scenario), assessing how the balance of RES capacity affects the occurrence of RES droughts.

@@

In this study, a RES drought event is defined as occurring when the 24-hour moving average of CF remains below a fixed threshold of 0.1 for a period of longer than 24 hours. The choice of this threshold is somewhat arbitrary, but aligns with similar studies on low renewable energy production \cite{kaspar2019drought, ohba2022drought, mayer2023drought} . By using a 24-hour moving average, fewer but longer-lasting events are captured compared to using the raw CF time series, which can be more sensitive to short-term fluctuations. A fixed threshold approach has been chosen in this study to enable consistent inter-comparison between datasets.

@@

The moving average approach smooths out short-term fluctuations, so that brief periods above the threshold do not interrupt an otherwise continuous low-CF period (Fig.~\ref{fig:find_res_droughts}). This means that a single hour above the threshold does not "break" a drought event if it is surrounded by prolonged low-generation hours. As a result, fewer but longer-lasting drought events are identified, which may better reflect real-world conditions where energy supply constraints persist over extended periods.

@@

The accuracy of the datasets used in this study are verified, before continuing to the analysis of RES droughts. For the verification process, we use time-varying values of installed capacity to account for changes in RES development over the verification period. This step allows us to assess how well the datasets represent the production of renewable energy by comparing them against observed data.

@@

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve}a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provide four associated smoothed power curves. The smoothing is done by using a Gaussian filter with different standard deviations that depend on the wind speed. A separate wind CF time series for Ireland was generated for each of the wind turbine power curves and smoothing levels. The performance of each CF time series was then calculated based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$  smoothing filter.

@@

The smoothing of the wind turbine power curve represents losses associated with each individual turbine, as well as losses such as wake effects between turbines, which are important when modelling wind energy on larger spatial scales. The histogram in Fig.~\ref{fig:power_curve}b shows that the C3S-E power curve tends to underestimate low CF values and overestimate higher ones, whereas the smoothed Atlite power curve more closely follows the recorded wind availability data from EirGrid.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_wind_contour}
	\caption{\textcolor{blue}{Save as png? Investigate vertical blobs near zero in (c)}Wind CF density plot of the observed EirGrid CF (vertical axes) and modelled (horizontal axes) CF data for the Atlite  (left panel), C3S-E G (central panel) and C3S-E N (right panel) models}
	\label{fig:wind_verification_contour}
\end{figure}

The effect of the difference between the power curves is also visible in Fig.~\ref{fig:wind_verification_contour}, which shows a density plot of wind CF values. The two C3S-E datasets are shown to overestimate the EirGrid CF, whereas the Atlite model is in good agreement with the EirGrid data. The skill scores presented in Table~\ref{tab:wind_skill_scores} show that Atlite performs better than the C3S-E datasets for all of the skill scores.

@@

The Atlite model allows the user to select certain PV panel characteristics. In this study we consider the three PV panels types available in the Atlite model. Following the same methodology as in the previous section, we compare the three available models using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, we selected the Breyer PV panel model \cite{beyer2004pv}, using the Kaneka Hybrid panel option.

@@

Fig.~\ref{fig:bar_number_events_verification_pv} shows the number of PV drought events in 2023 across different duration ranges. The figure reveals partial agreement between the three datasets and the EirGrid data, with consistent results observed for duration ranges of 1-2, 3-4, 7-8, and 8+ days. However, discrepancies appear in the other ranges, where the models diverge from the EirGrid data. The main challenge in validating PV data stems from the recent installation of a large share of Ireland’s PV capacity, leading to uncertainties in PV generation data and the actual generating capacity in the first few months after each farm is connected. With over 60\% of the total PV capacity installed in 2023, these data uncertainties significantly impact the ability to perform rigorous validation for PV drought events.

@@

In this section, we evaluate RES drought events under two different scenarios with fixed installed capacities: the existing scenario, with 89\% of wind capacity (6.7 GW) and 11\% of PV capacity (0.6 GW), and a balanced scenario, where wind capacity comprises 53\% (9 GW) and PV capacity increases to 47\% (8 GW). Both scenarios are driven by 45 years of ERA5 data. Using the RES drought identification process described in Section~\ref{sec:res_drought}, we first analyse wind and PV droughts separately before presenting the results for combined (wind + PV) RES droughts.

@@

The bottom panels show the combination of wind and PV under the  two capacity scenarios. In the existing RES installation scenario (Fig.~\ref{fig:boxplot_number_events}c), the identified RES droughts closely match those for wind alone, which is expected due to the dominance of installed wind capacity. In contrast, the balanced scenario (Fig.~\ref{fig:boxplot_number_events}d) shows a clear reduction in the number of drought events across all datasets and durations, with a decrease of the total number of events of 56\% for Atlite, 52\% for C3S-E G, and 50\% for C3S-E N. This reduction is attributed to the anti-correlation between wind and PV production throughout the year, a phenomenon well-documented in the literature \cite{kaspar2019drought}.

@@

The RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period refers to the estimated average interval between occurrences of events with a specified duration or intensity. Fig.~\ref{fig:return_periods} illustrates the return periods for varying RES drought durations, highlighting how often different drought lengths are likely to occur across the datasets. This analysis provides insight into the frequency and likelihood of prolonged low-generation periods, which is crucial for evaluating the potential impact of RES droughts on energy reliability and security of supply.

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves differently than the two C3S-E datasets. The Atlite results show a log-linear increase, but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets likely arises from how each model handles PV output near the threshold of 0.1 CF. Atlite remains slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.

For the existing installed capacity scenario (Fig.\ref{fig:return_periods}c), the return periods mirror those of wind (Fig.\ref{fig:return_periods}a), due to the low levels of installed PV. In the balanced scenario (Fig.~\ref{fig:return_periods}d), the return periods for RES droughts increase across all durations. For example, the return period for a five-day drought event, shown by the vertical dashed lines in Fig.~\ref{fig:return_periods}, extends from roughly six months for the existing scenario, to four years for the balanced scenario in the Atlite dataset, and from about fifteen months to around five years in the two C3S-E datasets. \textcolor{blue}{Shorten or remove: }The return periods for Atlite align more closely with those of C3S-E in the balanced scenario compared to the existing one. This convergence is due to the reduced proportion of wind capacity in the balanced scenario, which lessens the impact of wind-specific characteristics. Notably, there is a greater difference in how each model represents wind generation than PV, meaning the balanced scenario reduces these wind-related discrepancies, bringing the datasets closer together.

@@

Overall, the four panels indicate that regardless of the threshold applied, the Atlite dataset consistently identifies more events than the two C3S-E datasets, except for PV when the threshold is below 0.08. This finding suggests that Atlite, likely due to its more accurate representation of wind generation in Ireland, captures more RES drought events that might be missed with different modeling assumptions in the C3S-E datasets.
